An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-31 article/article.tex
finished paper

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (\pkg{Engine for likelihood-free
  inference})~\citet{1708.00707}. The implementation has been designed
to ensure extensibility. ROMC is an LFI framework; it defines a
sequence of algorithmic steps for approximating the posterior without
enforcing a specific algorithm for solving each step. Therefore, a
researcher may use ROMC as the backbone algorithm and develop novel
methods to solve each separate step. For being a ready-to-use
algorithm,~\citet{Ikonomov2019} proposed a default method for each
step, but this choice is by no means restrictive. We have designed our
software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
illustration and the evaluation of our implementation in three
steps. First, for securing that our implementation is accurate, we
test it against an artificial example with a tractable likelihood. The
artificial example also serves as a step-by-step guide for showcasing
how to use the various functionalities of our implementation. Second,
we use the second-order moving average (MA2) example from the
\pkg{ELFI} package, using as ground truth the samples obtained with
Rejection ABC~\citet{lintusaari2017} using a very high number of
samples. Finally, we present the execution times of ROMC, measuring
the speed-up achieved by the parallel version of the
implementation.

The code of the implementation is available at the official \pkg{ELFI} \href{https://github.com/elfi-dev/elfi}{repository}. Apart
from the examples presented in the paper, there are five \pkg{Google
  Colab} notebooks available online, with end-to-end examples
illustrating:
(i) \href{https://colab.research.google.com/drive/1lGRp0XrNfZ64NN0ASB_tYEKowXwlveDC?usp=sharing}{ROMC on a synthetic \(1D\) example},
(ii) \href{https://colab.research.google.com/drive/1Fof_WmCi1YizzSI_63aEsbLXsno5gSZ3?usp=sharing}{ROMC on a synthetic \(2D\) example},
(iii) \href{https://colab.research.google.com/drive/1nkdACQ370SSc0KB1bHv4sBRaxMlMqoNH?usp=sharing}{ROMC on the Moving Average example},
(iv) \href{https://colab.research.google.com/drive/1_jHVxPSH3XcNOORZJpLU0SPzs0PF8CQ5?usp=sharing}{how to extend ROMC with a Neural Network as a surrogate model},
(v) \href{https://colab.research.google.com/drive/1RzB-V1QueP1y1nyzv_VOqR1nVz3DUH3v?usp=sharing}{how to extend ROMC with a custom proposal region module}.

@@

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions \(\hat{\accregioni}\), which expands to three
mandatory tasks; (a) sample the nuisance variables
\(\ub_i \sim p(\ub)\) for defining the deterministic distance
functions \(d_i(\thetab)\) (Steps 3-5), (b) solve the optimization
problems for obtaining \(\thetab_i^*, d_i^*\) and keep the solutions
inside the threshold \(\epsilon\) (Steps 6-9), and (c) estimate the
bounding boxes \(\accregionihat\) to define uniform distributions
\(q_i\) on them (Step 10). Optionally, a surrogate model
\(\tilde{d}_i\) can be fitted for a faster inference phase (Step 11).

@@

After obtaining the optimal points \(\thetab^*_i\), we estimate the
proposal regions. Algorithm~\ref{alg:region_construction} describes
the line search approach for finding the region boundaries. The axes
of each bounding box \(\vb_m, m = \{1, \ldots, D\}\) are defined as
the directions with the highest curvature at \(\thetab_i^*\) computed
by the eigenvalues of the Hessian matrix \(\hessian_i\) of \(d_i\) at
\(\thetab_i\) (Step 1). Depending on the algorithm used in the
optimization step, we either use the real distance \(d_i\) or the
Gaussian Process approximation \(\hat{d}_i\). When the distance
function is the Euclidean distance (default choice), the Hessian
matrix can be also computed as \(\hessian_i = \jac_i^T\jac_i\), where
\(\jac_i\) is the Jacobian matrix of the summary statistics
\(\Phi(g(\thetab, \ub_i))\) at \(\thetab_i^*\). This approximation has
the computational advantage of using only first-order
derivatives. After defining the axes, we search for the bounding box
limits with a line step algorithm (Steps 2-21). The key idea is to
take long steps \(\eta\_\text{start}\) until crossing the boundary and
then take small steps backwards to find the exact boundary position.

@@

The results in this paper were obtained using \proglang{Python~3.7.9}
with the \pkg{ELFI~0.8.3} package. The experiments have been executed
in a single machine with an \proglang{Intel® Core™ i7-8750H} Processor
an with \proglang{Ubuntu 20.04} lts operating system.

>>>>

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (\pkg{Engine for likelihood-free
  inference})~\citet{1708.00707}. The implementation has been designed
to facilitate extensibility. ROMC is an LFI framework; it defines a
sequence of algorithmic steps for approximating the posterior without
enforcing a specific algorithm for each step. Therefore, a researcher
may use ROMC as the backbone method and apply novel algorithms to each
separate step. For being a ready-to-use LFI
method,~\citet{Ikonomov2019} propose a particular (default) algorithm
for each step, but this choice is by no means restrictive. We have
designed our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
illustration and the evaluation of our implementation in three
steps. First, for securing that our implementation is accurate, we
test it against an artificial example with a tractable likelihood. The
artificial example also serves as a step-by-step guide for showcasing
the various functionalities of our implementation. Second, we use the
second-order moving average (MA2) example~\citet{Marin2012} from the
\pkg{ELFI} package, using as ground truth the samples obtained with
Rejection ABC~\citet{lintusaari2017} using a very high number of
samples. Finally, we present the execution times of ROMC, measuring
the speed-up achieved by the parallel version of the implementation.

The code of the implementation is available at the official \pkg{ELFI}
\href{https://github.com/elfi-dev/elfi}{repository}. Apart from the
examples presented in the paper, there are five \pkg{Google
  Colab}~\citet{Bisong2019} notebooks available online, with
end-to-end examples illustrating how to: (i)
\href{https://colab.research.google.com/drive/1lGRp0XrNfZ64NN0ASB_tYEKowXwlveDC?usp=sharing}{use ROMC
  on a synthetic \(1D\) example}, (ii)
\href{https://colab.research.google.com/drive/1Fof_WmCi1YizzSI_63aEsbLXsno5gSZ3?usp=sharing}{use ROMC
  on a synthetic \(2D\) example}, (iii)
\href{https://colab.research.google.com/drive/1nkdACQ370SSc0KB1bHv4sBRaxMlMqoNH?usp=sharing}{use ROMC
  on the Moving Average example}, (iv)
\href{https://colab.research.google.com/drive/1_jHVxPSH3XcNOORZJpLU0SPzs0PF8CQ5?usp=sharing}{extend ROMC with a Neural Network as a surrogate model}, (v)
\href{https://colab.research.google.com/drive/1RzB-V1QueP1y1nyzv_VOqR1nVz3DUH3v?usp=sharing}{extend ROMC with a custom proposal region module}.

@@

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions \(\hat{\accregioni}\), which expands to three
mandatory tasks; (a) sample the nuisance variables
\(\ub_i \sim p(\ub)\) for defining the deterministic distance
functions \(d_i(\thetab)\) (steps 3-5), (b) solve the optimization
problems for obtaining \(\thetab_i^*, d_i^*\) and keep the solutions
inside the threshold \(\epsilon\) (steps 6-9), and (c) estimate the
bounding boxes \(\accregionihat\) to define uniform distributions
\(q_i\) on them (step 10). Optionally, a surrogate model
\(\tilde{d}_i\) can be fitted for a faster inference phase (step 11).

@@

After obtaining the optimal points \(\thetab^*_i\), we estimate the
proposal regions. Algorithm~\ref{alg:region_construction} describes
the line search approach for finding the region boundaries. The axes
of each bounding box \(\vb_m, m = \{1, \ldots, D\}\) are defined as
the directions with the highest curvature at \(\thetab_i^*\) computed
by the eigenvalues of the Hessian matrix \(\hessian_i\) of \(d_i\) at
\(\thetab_i\) (step 1). Depending on the algorithm used in the
optimization step, we either use the real distance \(d_i\) or the
Gaussian Process approximation \(\hat{d}_i\). When the distance
function is the Euclidean distance (default choice), the Hessian
matrix can be also computed as \(\hessian_i = \jac_i^T\jac_i\), where
\(\jac_i\) is the Jacobian matrix of the summary statistics
\(\Phi(g(\thetab, \ub_i))\) at \(\thetab_i^*\). This approximation has
the computational advantage of using only first-order
derivatives. After defining the axes, we search for the bounding box
limits with a line step algorithm (steps 2-21). The key idea is to
take long steps \(\eta\_\text{start}\) until crossing the boundary and
then take small steps backwards to find the exact boundary position.

@@

The results in this paper were obtained using \proglang{Python~3.9}
with the \pkg{ELFI~0.8.4} package. The experiments have been executed
in a single machine with an \proglang{Intel® Core™ i7-8750H} Processor
an with \proglang{Ubuntu 20.04} lts operating system.
